Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-05-29 manuscript.tex
Discuss implementation, add depletion chain plot

Using all incident neutron, decay, and fission product yield sublibrary files
from an ENDF-format library, such as ENDF/B-VIII.0, results in a \emph{full}
depletion chain that typically includes thousands of nuclides. Because the use
of a chain with $n$ nuclides will result in an $n\times n$ burnup matrix, there
is an incentive to reduce the number of nuclides in the depletion chain to
minimize both the size of the burnup matrix and the number of nuclides that are
tracked during a transport solve. The depletion chain class in OpenMC's Python
API includes a method for automatically reducing the size of the chain. Starting
with a set of nuclides that must be included, the method will recursively go
through all decay and reaction modes to determine what products may be produced.
Using this method on the ENDF/B-VII.1 library reduces the total number of
nuclides in the depletion chain from 3,820 nuclides to XXX.

In addition to the full depletion chain, a simplified depletion chain based on
the VERA depletion benchmark specification~\cite{kim2015casl} was created and
made available for users at \href{https://openmc.org}. This chain includes 223
nuclides and has adjusted branching ratios for some nuclides to preserve
accuracy of nuclide production rates.

@@

OpenMC does not currently allow energy-dependent capture branching ratios.
However, the depletion chain file does allow a reaction to be listed multiple
times with different branching ratios that result in different products.
Spectrum-averaged capture branching ratios have been computed in LWR and SFR
spectra and are available at \href{https://openmc.org/depletion-chains}. Within
the Python API, the class representing a depletion chain allows branching ratios
to be easily retrieved and changed. We note that while this treatment doesn't
accurately account for the change in the capture branching ratio with increasing
burnup (due to spectral changes), these changes are currently much smaller than
the differences in capture branching ratios between nuclear data evaluations.

@@

Different codes have adopted different approaches in how to handle the energy
dependence of FPYs. In ORIGEN, the average neutron energy causing fission is
calculated, and FPYs are interpolated using this average
energy~\citep{gauld2011nt}. In MCNP, the fission rate is tallied over a few
energy groups (thermal, fast, and high energy), and the energy group that
contains the majority of fissions is used to select appropriate
FPYs~\citep{fensin2008nt}. In Serpent, a linear interpolation factor is scored
based on the neutron energy causing fission in relationship to the energy points
at which FPY are tabulated; the interpolation factor is then used to calculate
an effective FPY~\citep{kunchev2019mc}.

@@

Class organization with Operator and Integrators separate. Use of shared library
and in-memory handling through Python API. This allows time loading nuclear data
to be amortized since it only needs to be done once.

Implementation of depletion in Python. Relying on scipy sparse solve for the
CRAM system. Parallelization through Python multiprocessing pool.

>>>>

Using all incident neutron, decay, and fission product yield sublibrary files
from an ENDF-format library, such as ENDF/B-VIII.0, results in a \emph{full}
depletion chain that typically includes thousands of nuclides. Because the use
of a chain with $n$ nuclides will result in an $n\times n$ burnup matrix, there
is an incentive to reduce the number of nuclides in the depletion chain to
minimize both the size of the burnup matrix and the number of nuclides that are
tracked during a transport solve. The depletion chain class in OpenMC's Python
API includes a method for automatically reducing the size of the chain. Starting
with a set of nuclides that must be included, the method will recursively go
through all decay and reaction modes to determine what products may be produced.
Using this method on the ENDF/B-VII.1 library starting from a list of all stable
or very long lived (half-life greater than $10^{15}$ sec) nuclides reduces the
total number of nuclides in the depletion chain from 3,820 to 1,693.

In addition to the full depletion chain, a simplified depletion chain based on
the VERA depletion benchmark specification~\cite{kim2015casl} was created and
made available for users at \url{https://openmc.org}. This chain includes 223
nuclides and has adjusted branching ratios for some nuclides to preserve
accuracy of nuclide production rates. \Cref{fig:chains} shows which nuclides are
included in each of the depletion chains.
\begin{figure}
  \centering
  \includegraphics[width=4in]{figures/chains.pdf}
  \caption{Table of nuclides showing which nuclides are included in each of the
    depletion chains based on ENDF/B-VII.1.}
  \label{fig:chains}
\end{figure}

@@

OpenMC does not currently allow energy-dependent capture branching ratios.
However, the depletion chain file does allow a reaction to be listed multiple
times with different branching ratios that result in different products.
Spectrum-averaged capture branching ratios have been computed in LWR and SFR
spectra and are available at \url{https://openmc.org/depletion-chains}. Within
the Python API, the class representing a depletion chain allows branching ratios
to be easily retrieved and changed. We note that while this treatment doesn't
accurately account for the change in the capture branching ratio with increasing
burnup (due to spectral changes), these changes are currently much smaller than
the differences in capture branching ratios between nuclear data evaluations.

@@

Different codes have adopted different approaches in how to handle the energy
dependence of FPYs. In ORIGEN, the average neutron energy causing fission is
calculated, and FPYs are interpolated using this average
energy~\citep{gauld2011nt}. In MCNP, the fission rate is tallied over a few
energy groups (thermal, fast, and high energy), and the energy group that
contains the majority of fissions is used to select appropriate
FPYs~\citep{fensin2008nt,fensin2011lanl}. In Serpent, a linear interpolation
factor is scored based on the neutron energy causing fission in relationship to
the energy points at which FPY are tabulated; the interpolation factor is then
used to calculate an effective FPY~\citep{kunchev2019mc}.

@@

OpenMC's core particle transport solver is written in C++ and is built as a
single executable. Most of the user-facing code, however, is written in Python.
Generally, to run OpenMC, a user writes a Python script that utilizes classes
and functions defined as part of an application programming interface (API) to
generate XML input files that are then read by the executable. On top of this,
OpenMC is also built as a shared library with a C++ API that can be called by
external programs. OpenMC's Python API includes bindings to the C++ API; this
enables a single run to be broken up into individual components. For example, a
function call from the Python bindings to the C++ API can initialize the solver
(reading inputs, loading cross sections, etc.) without actually performing a
simulation. A separate function call can run the transport simulation. Further
function calls allow a power user to obtain information about tally results
directly in memory from the C++ API.

OpenMC's depletion capability is implemented entirely in Python as the
\texttt{openmc.deplete} module. This module relies on the Python bindings to the
C++ API to initialize OpenMC, create tallies necessary for depletion, run
transport simulations, and obtain tally results. This approach has two primary
advantages. First, the time necessary to load cross sections is amortized
because it only has to be performed once, as opposed to once per transport
simulation. Second, using the Python bindings to the C++ API allows results to
be read directly in memory rather than using filesystem I/O, which helps to
minimizing overhead, particularly for large simulations.

The \texttt{openmc.deplete} module is designed with three key abstract classes
that represent the major components of the depletion solver. Each of the
abstract classes themselves defines an interface, that is, a set of methods that
a subclass must implement. First, the abstract \texttt{TransportOperator} class
represents a particle transport solver. Subclasses of \texttt{TransportOperator}
are required to implement a method that takes a list of nuclide densities to be
used in the problem and a system power and returns the $k$-eigenvalue and the
reaction rates resulting from the transport operator. Currently, there is a
single subclass of \texttt{TransportOperator} that interfaces with the Python
bindings to OpenMC's C++ API to call a tranport solve and obtain reaction rates.
Once reaction rates are obtained, a separate depletion chain class (that is not
abstract) uses them to construct the burnup matrix, $\vect{A}$, based on the
data present in the depletion chain.

The abstract \texttt{DepSystemSolver} class represents a solver for
\cref{eq:depletion-matrix} when $\vect{A}$ is constant (a solver that can
evaluate \cref{eq:constant-A}). Subclasses of \texttt{DepSystemSolver} are
required to implement a method that takes the burnup matrix $\vect{A}$, an
initial density vector $\vect{n}_0$, and a timestep size, and returns the
resulting nuclide densities at the end of the step. The only subclass of
\texttt{DepSystemSolver} at present is \texttt{IPFCramSolver}, which implements
the IPF form of CRAM to evaluate \cref{eq:constant-A}. Because the CRAM method
requires solving multiple sparse linear systems, this class relies on the
third-party \texttt{scipy.sparse} package for performing a sparse linear solve
(REF).

The abstract \texttt{Integrator} class represents a time integration method.
This class effectively orchestrates the coupled transport-depletion simulation.
Subclasses of \texttt{Integrator} are required to implement a method that takes
an initial vector of nuclide densities, reaction rates returned from an
operator, a time interval, and a power, and returns the resulting nuclide
densities at the end of the step. Thus, \texttt{Integrator} utilizes
\texttt{TransportOperator} in order to obtain reaction rates given $\vect{n}$,
and once the burnup matrix is constructed, it utilizes \texttt{DepSystemSolver}
to deplete with a constant $\vect{A}$ as many times as is necessary. One example
of a subclass of \texttt{Integrator} is \texttt{CECMIntegrator}, which
implements the CE/CM method.

The depletion system in OpenMC is designed to be modular. Although the only
transport operator that is implemented is the OpenMC transport solver itself,
the rest of the infrastructure (time integration methods, CRAM solver) does not
rely on anything specific to OpenMC's transport solver, so one could in
principle implement a class that utilizes a different transport solver (whether
that be another Monte Carlo code or a deterministic code). Similarly, although
\texttt{IPFCramSolver} is the only depletion system solver implemented, one
could replace it with a different solver, for example, a solver based on
backward differentiation formulas.

Lastly, we note that when multiple depletable materials are present in a problem
(thereby requiring multiple independent solutions of
\cref{eq:depletion-matrix}), the work is fully parallelized over the available
processes/threads. The third-party \texttt{mpi4py} Python package is used to
distribute the material compositions among the participating MPI processes.
Then, within each process, Python's \texttt{multiprocessing} standard-library
module is used to distribute materials among threads within a single MPI
process. This matches the distributed- and shared-memory parallelization scheme
used in OpenMC's transport solver, which relies on MPI and OpenMP.
